---
title: "Small Worlds and Large Worlds"
author: "schifferl"
date: "February 21, 2018"
output:
  html_document:
    fig_height: 7
    fig_retina: NULL
    code_folding: "hide"
css: css/style.css
editor_options: 
  chunk_output_type: console
---

## `knitr` Options

```{r}
knitr::opts_chunk$set(fig.align = "center")
knitr::opts_chunk$set(results = "hold")
```

## `R` Packages

```{r, message=FALSE}
# TODO
```

## Pratice

### Easy.

2E1. Which of the expressions below correspond to the statement: the probability
of rain on Monday?

  (1) Pr(rain)
  (2) Pr(rain|Monday)
  (3) Pr(Monday|rain)
  (4) Pr(rain, Monday) / Pr(Monday)
  
>$$P(rain\ |\ Monday)$$
>*Assuming independence between the day of the week and rain, also:*
>$$P(rain, Monday)\ /\ P(Monday)$$
>*This is because:*
>$$P(rain)=P(rain\ |\ Monday)$$

2E2. Which of the following statements corresponds to the expression: 
Pr(Monday|rain)?

  (1) The probability of rain on Monday.
  (2) The probability of rain, given that it is Monday.
  (3) The probability that it is Monday, given that it is raining.
  (4) The probability that it is Monday and that it is raining.
  
>*The probability that it is Monday, given that it is raining.*

2E3. Which of the expressions below correspond to the statement: the probability
that it is Monday, given that it is raining?

  (1) Pr(Monday|rain)
  (2) Pr(rain|Monday)
  (3) Pr(rain|Monday) Pr(Monday)
  (4) Pr(rain|Monday) Pr(Monday) / Pr(rain)
  (5) Pr(Monday|rain) Pr(rain) / Pr(Monday)
  
>$$P(Monday\ |\ rain)$$

2E4. The Bayesian statistician Bruno de Finetti (1906–1985) began his book on
probability theory with the declaration: "PROBABILITY DOES NOT EXIST." The
capitals appeared in the original, so I imagine de Finetti wanted us to shout
this statement. What he meant is that probability is a device for describing
uncertainty from the perspective of an observer with limited knowledge; it has
no objective reality. Discuss the globe tossing example from the chapter, in
light of this statement. What does it mean to say "the probability of water is
0.7"?

>*The statement that "PROBABILITY DOES NOT EXIST"...*

### Medium.

2M1. Recall the globe tossing model from the chapter. Compute and plot the grid
approximate posterior distribution for each of the following sets of
observations. In each case, assume a uniform prior for p.

  (1) W, W, W
  
      > ```{r}
        # define grid
        p_grid <- seq(from = 0, to = 1, length.out = 100)
        
        # define prior
        prior <- rep(1, 100)
        
        # compute likelihood at each value in grid
        likelihood <- dbinom(3, size = 3, prob = p_grid)
        
        # compute product of likelihood and prior
        unstandardized_posterior <- likelihood * prior
        
        # standardize the posterior, so it sums to 1
        posterior <- unstandardized_posterior / sum(unstandardized_posterior)
        
        # plot the grid approximate posterior probability of water
        plot(p_grid, posterior, type = "b", main = "Grid Approximate Posterior",
             xlab = "Probability of Water", ylab = "Posterior Probability")
        ```
  
  (2) W, W, W, L
  
      > ```{r}
        # define grid
        p_grid <- seq(from = 0, to = 1, length.out = 100)
        
        # define prior
        prior <- rep(1, 100)
        
        # compute likelihood at each value in grid
        likelihood <- dbinom(3, size = 4, prob = p_grid)
        
        # compute product of likelihood and prior
        unstandardized_posterior <- likelihood * prior
        
        # standardize the posterior, so it sums to 1
        posterior <- unstandardized_posterior / sum(unstandardized_posterior)
        
        # plot the grid approximate posterior probability of water
        plot(p_grid, posterior, type = "b", main = "Grid Approximate Posterior",
             xlab = "Probability of Water", ylab = "Posterior Probability")
        ```
  
  (3) L, W, W, L, W, W, W
  
      > ```{r}
        # define grid
        p_grid <- seq(from = 0, to = 1, length.out = 100)
        
        # define prior
        prior <- rep(1, 100)
        
        # compute likelihood at each value in grid
        likelihood <- dbinom(5, size = 7, prob = p_grid)
        
        # compute product of likelihood and prior
        unstandardized_posterior <- likelihood * prior
        
        # standardize the posterior, so it sums to 1
        posterior <- unstandardized_posterior / sum(unstandardized_posterior)
        
        # plot the grid approximate posterior probability of water
        plot(p_grid, posterior, type = "b", main = "Grid Approximate Posterior",
             xlab = "Probability of Water", ylab = "Posterior Probability")
        ```
  
2M2. Now assume a prior for p that is equal to zero when p < 0.5 and is a 
positive constant when p ≥ 0.5. Again compute and plot the grid approximate 
posterior distribution for each of the sets of observations in the problem just
above.

> ```{r}
  # TODO
  ```

2M3. Suppose there are two globes, one for Earth and one for Mars. The Earth 
globe is 70% covered in water. The Mars globe is 100% land. Further suppose that
one of these globes—you don’t know which—was tossed in the air and produced a
"land" observation. Assume that each globe was equally likely to be tossed. Show
that the posterior probability that the globe was the Earth, conditional on
seeing "land" (Pr(Earth|land)), is 0.23.

> ```{r}
  # TODO
  ```

2M4. Suppose you have a deck with only three cards. Each card has two sides, and
each side is either black or white. One card has two black sides. The second
card has one black and one white side. The third card has two white sides. Now
suppose all three cards are placed in a bag and shuffled. Someone reaches into
the bag and pulls out a card and places it flat on a table. A black side is
shown facing up, but you don’t know the color of the side facing down. Show that
the probability that the other side is also black is 2/3. Use the counting
method (Section 2 of the chapter) to approach this problem. This means counting
up the ways that each card could produce the observed data (a black side facing
up on the table).

> ```{r}
  # TODO
  ```

2M5. Now suppose there are four cards: B/B, B/W, W/W, and another B/B. Again
suppose a card is drawn from the bag and a black side appears face up. Again
calculate the probability that the other side is black.

> ```{r}
  # TODO
  ```

2M6. Imagine that black ink is heavy, and so cards with black sides are heavier
than cards with white sides. As a result, it’s less likely that a card with
black sides is pulled from the bag. So again assume there are three cards: B/B,
B/W, and W/W. After experimenting a number of times, you conclude that for every
way to pull the B/B card from the bag, there are 2 ways to pull the B/W card and
3 ways to pull the W/W card. Again suppose that a card is pulled and a black
side appears face up. Show that the probability the other side is black is now
0.5. Use the counting method, as before.

> ```{r}
  # TODO
  ```

2M7. Assume again the original card problem, with a single card showing a black
side face up. Before looking at the other side, we draw another card from the
bag and lay it face up on the table. The face that is shown on the new card is
white. Show that the probability that the first card, the one showing a black
side, has black on its other side is now 0.75. Use the counting method, if you
can. Hint: Treat this like the sequence of globe tosses, counting all the ways
to see each observation, for each possible first card.

> ```{r}
  # TODO
  ```

### Hard.

2H1. Suppose there are two species of panda bear. Both are equally common in the
wild and live in the same places. They look exactly alike and eat the same food,
and there is yet no genetic assay capable of telling them apart. They differ
however in their family sizes. Species A gives birth to twins 10% of the time,
otherwise birthing a single infant. Species B births twins 20% of the time,
otherwise birthing singleton infants. Assume these numbers are known with
certainty, from many years of field research.

Now suppose you are managing a captive panda breeding program. You have a new
female panda of unknown species, and she has just given birth to twins. What is
the probability that her next birth will also be twins?

> ```{r}
  # TODO
  ```

2H2. Recall all the facts from the problem above. Now compute the probability
that the panda we have is from species A, assuming we have observed only the
first birth and that it was twins.

> ```{r}
  # TODO
  ```

2H3. Continuing on from the previous problem, suppose the same panda mother has
a second birth and that it is not twins, but a singleton infant. Compute the
posterior probability that this panda is species A.

> ```{r}
  # TODO
  ```

2H4. A common boast of Bayesian statisticians is that Bayesian inference makes
it easy to use all of the data, even if the data are of different types.

So suppose now that a veterinarian comes along who has a new genetic test that
she claims can identify the species of our mother panda. But the test, like all
tests, is imperfect. This is the information you have about the test:

- The probability it correctly identifies a species A panda is 0.80.
- The probability it correctly identifies a species B panda is 0.65.

The vet administers the test to your panda and tells you that the test is
positive for species A. First ignore your previous information from the births
and compute the posterior probability that your panda is species A. Then redo 
your calculation, now using the birth data as well.

> ```{r}
  # TODO
  ```

> ```{r}
  # TODO
  ```
