---
title: "Chapter 3H Exercises"
author: "Sal Leggio"
date: "March 2, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r , echo = TRUE}

library(rmarkdown)
library(plotly)
library(tidyverse)
library(rethinking)

rm(list=ls(all=TRUE))
data(homeworkch3)
```
3H1
===

## Using grid approximation, compute the posterior distribution for the probability of a birth being a boy. Assume a uniform prior probability. Which parameter value maximizes the posterior probability?

## what should we use for likelihood?

- likelihood is  L(theta | x) = P (X=x | theta) 
- function of the parameter, not of the data
- gives probability of the data if the parameter = theta

For this problem, it'a a binomial distribution.
How many boys should we expect in families with two children?
What is P(boy)?

Two Reasonable Answers:
- (1) common sense says P(boy) = 0.5
- (2) demography says global sex ratio at birth = 1.07 = (number of boys) / (number of girls). This implies P(boy) = (1.07)/(2.07) = .517

Less Reasonable
- (3) our data says 111 boys out of 200 children, so P(boy) = 0.555. This is much higher than the global average, and is probably due to sampling error, but that's our data.


First we'll look at option (1)

```{r, echo=TRUE}
# define array of parameter values
sex_grid <- seq( from = 0 , to = 1 , length.out = 100 ) 


# define prior - initially uniform 
prior <- rep( 1 , 100)

#compute likelihood at each point 
likelihood <- dbinom( 1, size=2 , prob = sex_grid) # chance of 1 boy in a 2 child family

#compute posterior
postr   <- likelihood * prior
posterior <- postr / sum(postr) 

# plot
plot (sex_grid , posterior , type = "b" , 
      xlab = "probability of boy" , ylab = "posterior probability")
mtext("100 pts")
```
## what parameter value maximizes posterior probability?

```{r , echo=TRUE}
which.max(posterior)
sex_grid[which.max(posterior)]  ## sex_grid is an array of parameter values
```
So our model seems to lean towards Option 1, so far.



3H2
===

## Using the sample function, draw 10,000 random parameter values from the posterior distribution. Calculate 50%, 89%, 97% highest posterior density intervals.

```{r, echo=TRUE}
# sampleSEX is a sample of the parameters from sex_grid, using the posterior probabilities
sampleSEX <- sample (sex_grid , prob = posterior , size = 1e4 , replace = TRUE ) 
HPDI( sampleSEX , prob = 0.50)
HPDI( sampleSEX , prob = 0.89)
HPDI( sampleSEX , prob = 0.97)
```


3H3 Part A
=========

## Use rbinom to simulate 10,000 replicates of 200 births. You should end up with 10,000 numbers, each one a count of boys out of 200 births. Compare the distribution of predicted numbers to the actual count in the data (111 boys out of 200).

```{r , echo=TRUE}

# I am guessing that we use all possible values of the parameter to do this
# w_boy is the number of boys in each sample of 200 kids
w_boy <- rbinom(1e4 , size = 200 , prob =  sampleSEX)

hist(w_boy , xlab = "No. of Boys" , main = "w_boy" ,
     border = "blue" , col = "green")

boys <- birth1 + birth2
hist(boys, xlab = "No. of Boys" , main = "Data",
     border = "blue" , col = "salmon")


```
