---
title: "Chapter 3H Exercises"
author: "Sal Leggio"
date: "March 2, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r , echo = TRUE}

library(rmarkdown)
library(plotly)
library(tidyverse)
library(rethinking)

rm(list=ls(all=TRUE))
data(homeworkch3)
```
3H1
===

## Using grid approximation, compute the posterior distribution for the probability of a birth being a boy. Assume a uniform prior probability. Which parameter value maximizes the posterior probability?

## what should we use for likelihood?

- likelihood is  L(theta | x) = P (X=x | theta) 
- function of the parameter, not of the data
- gives probability of the data if the parameter = theta

For this problem, it'a a binomial distribution.
How many boys should we expect in families with two children?
What is P(boy)?

Two Reasonable Answers:
- (1) common sense says P(boy) = 0.5
- (2) demography says global sex ratio at birth = 1.07 = (number of boys) / (number of girls). This implies P(boy) = (1.07)/(2.07) = .517

Less Reasonable
- (3) our data says 111 boys out of 200 children, so P(boy) = 0.555. This is much higher than the global average, and is probably due to sampling error, but that's our data.


First we'll look at option (1)

```{r, echo=TRUE}
# define array of parameter values
sex_grid <- seq( from = 0 , to = 1 , length.out = 100 ) 


# define prior - initially uniform 
prior <- rep( .01 , 100)

# compute likelihood at each point 
# using the data, 111 boys out of 200
# what is the probabiity of getting this result given various parameters?
likelihood <- dbinom( 111, size=200 , prob = sex_grid) # chance of 1 boy in a 2 child family

#compute posterior
postr   <- likelihood * prior
posterior <- postr / sum(postr)

# next line looks wrong, you are already multiplying normalized probabilities
#posterior <- postr / sum(postr) 

# plot
plot (sex_grid , posterior , type = "b" , 
      xlab = "probability of boy" , ylab = "posterior probability" ,
      col = "steelblue")
mtext("100 pts")
```
## what parameter value maximizes posterior probability?

```{r , echo=TRUE}
which.max(posterior)
sex_grid[which.max(posterior)]  ## sex_grid is an array of parameter values
```


3H2
===

## Using the sample function, draw 10,000 random parameter values from the posterior distribution. Calculate 50%, 89%, 97% highest posterior density intervals.

```{r, echo=TRUE}
# sampleSEX is a sample of the parameters from sex_grid, using the posterior probabilities
sampleSEX <- sample (sex_grid , prob = posterior , size = 1e4 , replace = TRUE ) 
HPDI( sampleSEX , prob = 0.50)
HPDI( sampleSEX , prob = 0.89)
HPDI( sampleSEX , prob = 0.97)
```


3H3 
=========

## Use rbinom to simulate 10,000 replicates of 200 births. You should end up with 10,000 numbers, each one a count of boys out of 200 births. Compare the distribution of predicted numbers to the actual count in the data (111 boys out of 200).

```{r , echo=TRUE}

# I am guessing that we use all possible values of the parameter to do this
# w_boy is the number of boys in each sample of 200 kids
samples1  <- sample(sex_grid, 1e4, replace = TRUE , prob = posterior)
w_boy <- rbinom(1e4 , size = 200 , prob = samples1)

hist(w_boy , xlab = "No. of Boys" , main = "w_boy" ,
     border = "blue" , col = "green")
lines (dens (w_boy , col = "red") )


boys <- birth1 + birth2
hist(boys, xlab = "No. of Boys" , main = "Data",
     border = "blue" , col = "salmon")


w_boy2 <- w_boy / 100


summary(w_boy2)
summary(boys)

## the distribution of the simulated data is different from the observed data

```

3H4
===

## Compare 10,000 counts of boys from 100 simulated first borns, to the number of boys in birth1

```{r, echo= TRUE}
sum(birth1)  ## 51 of the first born are male

w_birth1<- rbinom(1e4 , size = 100 , prob =  0.51)
summary(w_birth1)
## note that mean and median are 51, same as the number of boys in birth1


hist(w_birth1, xlab = "No. of Boys" , main = "Simulate First Born Males",
     border ="blue" , col = "salmon")

```

3H5
===

## The model assumes that sex of first and 2nd births are independent. To check this assumption, focus now on second births that followed female first borns (birth1 = 0).Compare 10,000 simulated counts of boys to only those second births that followed girls. 

## To do this correctly, you need to count the number of first born who were girls and simulate that many births, 10,000 times. 

Isn't this the same distribution as before? 49 girls in a hundred means 51 boys out of a hundred. Just subtract the boy counts form 100 to get girl counts.

   
## Compare the actual count of boys in the simulation to the observed counts of boys following girls

``` {r, echo = TRUE}
second_births <- tibble(birth1, birth2)
second_births <- filter (second_births, birth1 == 0)
   ## second births now has 49 observations, corresponding to the 49 first born girls

summary(second_births$birth2)


```
##  80% of the second births following the first birth of a girl are boys
