---
title: "Monte Carlo Estimation"
author: "Philipp Grafendorfer"
date: "25th of April 2018"
output: revealjs::revealjs_presentation
  reveal_plugins: ["notes"]
  self_contained: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
)
library(DiagrammeR)
library(png)
library(grid)
library(rethinking)
library(psych)
library(revealjs)

data(rugged)
d <- rugged
d$log_gdp <- log(d$rgdppc_2000)
dd <- d[ complete.cases(d$rgdppc_2000) , ]
dd.trim <- dd[ , c("log_gdp", "rugged", "cont_africa")]
```

## Overview

- Introduction and structure
- The Metropolis Algorithm
- Hamilton Monte Carlo
- Exercises and applications
- Bad sampling and how to tam a wild chain
- Compare different models

Note:
This is my *note*.


## Motivation and Introduction

### Stanislaw Marcin Ulam (1909-1984)

<div class="columns-2">

- stan programming language
- student of Banach
- early development of nuclear weapons
- main contributor to the Collatz problem

```{r fig.height=3,echo=FALSE, fig.align='center'}
img <- readPNG("img/Ulam_5.png")
 grid.raster(img)
```
</div>

----

### Nicholas Metropolis (1915-1999) 

<div class="columns-2">

- physicist
- developed the Monte Carlo Method with Ulam and von Neumann in Los Alamos
- inventor of simulated annealing

```{r fig.height=3,echo=FALSE, fig.align='center'}
img <- readPNG("img/metropolis_0.png")
 grid.raster(img)
```
</div>

----

### Edward Teller (1908-2003)

<div class="columns-2">

- "father" of hydrogen bomb
- "Dr. Strangelove"
- Manhattan Project

```{r fig.height=3,echo=FALSE, fig.align='center'}
img <- readPNG("img/edward_teller.png")
 grid.raster(img)
```
</div>

----

### The Journey of King Markov (through a discrete parameter space)

```{r fig.height=6,echo=FALSE, fig.align='center'}
img <- readPNG("img/king_markov.png")
 grid.raster(img)
```

----

### MCMC 

> - Sampling of (in general) non- normal posteriors
> - can be very efficient
> - output is a sample of posteriors
> - a byproduct of MCMC is a model fit but we are not (primarily) interested in that

----

### Structural Overview

```{r myplot,echo=FALSE,fig.width=10}
grViz("
      digraph boxes_and_circles {
      
      node [shape = circle]
      MCMC; Metropolis; 'Metropolis Hastings'; 'Hamilton MC'; 'Gibbs Sampling'; BUGS; JAGS

      node [shape = box
            fontname = Helvetica
            penwidth = 2.0]
      'symmetric proposal'; 'asymmetric proposal'; 'high cost, high efficiency'; 'continous parameters';
        'full sweep'; 'conjugate pairs/priors'; 'adaptive proposal'
      
      edge [ arrowhead = NULL]
      MCMC -> 'Metropolis' [label = ' algorithm'];
      MCMC -> 'Metropolis Hastings' [label = ' algorithm'];
      Metropolis -> 'symmetric proposal';
      'Metropolis Hastings' -> 'asymmetric proposal';
      'Metropolis Hastings' -> 'Gibbs Sampling' [label = ' technique'];
      'Metropolis Hastings' -> 'Hamilton MC' [label = ' technique'];
      'Gibbs Sampling' -> 'adaptive proposal';
      'Gibbs Sampling' -> BUGS [label = ' software'];
      'Gibbs Sampling' -> 'conjugate pairs/priors';
      'Gibbs Sampling' -> JAGS;
      'Hamilton MC' -> 'full sweep';
      'Hamilton MC' -> 'continous parameters';
      'Hamilton MC' -> 'high cost, high efficiency'
      
      }
      ")
```

## Markov Chain Monte Carlo

```{r fig.width=10,echo=FALSE, fig.cap="David Kipping, Sagan Workshop 2016"}
img <- readPNG("img/detail_view_mcmc_kippling.png")
 grid.raster(img)
```

----

```{r fig.width=10,echo=FALSE, fig.cap="David Kipping, Sagan Workshop 2016"}
img <- readPNG("img/reason_for_mcmc.png")
 grid.raster(img)
```

## The Metropolis Algorithm

```{r fig.width=10,echo=FALSE, fig.cap="David Kipping, Sagan Workshop 2016"}
img <- readPNG("img/metro_alg_1.png")
 grid.raster(img)
```

----

```{r fig.width=10,echo=FALSE, fig.cap="David Kipping, Sagan Workshop 2016"}
img <- readPNG("img/metro_alg_2.png")
 grid.raster(img)
```

----

```{r fig.width=10,echo=FALSE, fig.cap="David Kipping, Sagan Workshop 2016"}
img <- readPNG("img/metro_alg_3.png")
 grid.raster(img)
```

----

```{r fig.width=10,echo=FALSE, fig.cap="David Kipping, Sagan Workshop 2016"}
img <- readPNG("img/metro_alg_4.png")
 grid.raster(img)
```

----

```{r fig.width=10,echo=FALSE, fig.cap="David Kipping, Sagan Workshop 2016"}
img <- readPNG("img/metro_alg_5.png")
 grid.raster(img)
```

----

```{r fig.width=10,echo=FALSE, fig.cap="David Kipping, Sagan Workshop 2016"}
img <- readPNG("img/metro_alg_6.png")
 grid.raster(img)
```

----

```{r fig.width=10,echo=FALSE, fig.cap="David Kipping, Sagan Workshop 2016"}
img <- readPNG("img/metro_alg_7.png")
 grid.raster(img)
```

## Hamilton Monte Carlo (HMC)

*"It appears to be a quite general principle that, whenever there is a randomized*
*way of doing something, then there is a nonrandomized way that delivers*
*better performance but requires more thought."* —E. T. Jaynes

- HMC is less random
- only continous parameters
- needs hyperparameter tuning with stan
- works well with a big number of parameters
- sweeps over whole parameter space
- easy Parallelization
- from now on only HMC

## Exercise 8M1

*Re-estimate the terrain ruggedness model from the chapter, but now using a uniform prior and*
*an exponential prior for the standard deviation, sigma. The uniform prior should be dunif(0,10)*
*and the exponential should be dexp(1). Do the different priors have any detectible influence on the*
*posterior distribution?*

```{r eval=T, include=F}
# put a uniform prior on sigma
m8.1.unif <- readRDS("model/m8.1.unif.rds")
```

```{r eval=T, include=F}
# put an exponential prior on sigma
m8.1.exp <- readRDS("model/m8.1.exp.rds")
```

```{r fig.height=4, fig.align='center'}
# visualize each prior
# a legend can be added
curve(dcauchy(x, 0, 2), from = 0, to = 10, xlab = "sigma", ylab = "Prior Density", ylim = c(0, 1))
curve(dunif(x, 0, 10), from = 0, to = 10, add = TRUE, col = "blue")
curve(dexp(x, 1), from = 0, to = 10, add = TRUE, col = "red")
legend(6, 0.8, legend=c("Cauchy", "Uniform", "Exponential"), col=c("black", "blue", "red"), lty=1:1, cex=0.8, title = "Prior Distribution")
```

----

### Sampling the posterior distribution

```{r message=FALSE, warning=FALSE, fig.height=4, fig.align='center'}
# plot the posterior for sigma for each model
sigma_unif <- extract.samples(m8.1.unif,pars="sigma")
sigma_exp <- extract.samples(m8.1.exp,pars="sigma")
dens(sigma_unif[[1]], xlab="sigma", xlim=c(0.5,1.5), col="red")
dens(sigma_exp[[1]], add=TRUE, col="blue")
legend(1.2, 6, legend=c("Uniform", "Exponential"), col=c("red", "blue"), lty=1:1, cex=0.8, title = "Posterior Distribution")
```

----

### Pair Plot Posterior Distributions

```{r fig.cap="Posterior with Uniform Prior for Sigma", fig.align='center', fig.height=5, fig.width=10}
post.unif <- extract.samples(m8.1.unif)
pairs.panels(as.data.frame(post.unif))
```

---- 
 
### Pair Plot Posterior Distributions

```{r fig.cap="Posterior with Exponential Prior for Sigma", fig.align='center', fig.height=5, fig.width=10}
post.exp <- extract.samples(m8.1.exp)
pairs.panels(as.data.frame(post.exp))
```

----

### The precis function delivers

```{r}
precis(m8.1.exp)
```

- standard parameter for iterations: 2000
- standard parameter for warmup: 1000
- **n_eff** gives an estimated number of effective samples from the posterior respectively
- **Rhat** is the *Gelman Rubicon Convergence Diagnostic* for the markov chain (should be close to 1)
- The intervall is a Maximum A Posteriori (MAP) intervall

----

### Model summary

```{r}
show(m8.1.exp)
```

----

### Traceplot

Long warmup phase.

```{r fig.align='center', fig.height=4, fig.width=10}
plot(m8.1.exp)
```

## Exercise 8M2

*The Cauchy and exponential priors from the terrain ruggedness model are very weak. They*
*can be made more informative by reducing their scale. Compare the dcauchy and dexp priors for*
*progressively smaller values of the scaling parameter. As these priors become stronger, how does each*
*influence the posterior distribution?*

```{r}
m8.2.cauchy.10 <- readRDS("model/m8.2.cauchy.10.rds")

m8.2.cauchy.1 <- readRDS("model/m8.2.cauchy.1.rds")

m8.2.cauchy.point.1 <- readRDS("model/m8.2.cauchy.point.1.rds")
```

```{r}
m8.2.exp.10 <- readRDS("model/m8.2.exp.10.rds")

m8.2.exp.1 <- readRDS("model/m8.2.exp.1.rds")

m8.2.exp.point.1 <- readRDS("model/m8.2.exp.point.1.rds")
```

```{r fig.align='center', fig.width=10, fig.cap='Decreasing Scaling Parameter for Cauchy Distribution'}
curve(dcauchy(x, 0, 10), from = 0, to = 4, xlab = "sigma", ylab = "Prior Density", ylim = c(0, 1))
curve(dcauchy(x, 0, 1), from = 0, to = 4, add = TRUE, col = "blue")
curve(dcauchy(x, 0, 0.1), from = 0, to = 4, add = TRUE, col = "red")
legend(6, 0.8, legend=c("10", "1", "0.1"), col=c("black", "blue", "red"), lty=1:1, cex=0.8, title = "Prior Distribution")
```

----

### Decreasing Scaling Parameter for Exponential Distribution

```{r fig.align='center', fig.width=10}
curve(dexp(x, 10), from = 0, to = 10, xlab = "sigma", ylab = "Prior Density", ylim = c(0, 1))
curve(dexp(x, 1), from = 0, to = 10, add = TRUE, col = "blue")
curve(dexp(x, 0.1), from = 0, to = 10, add = TRUE, col = "red")
legend(6, 0.8, legend=c("10", "1", "0.1"), col=c("black", "blue", "red"), lty=1:1, cex=0.8, title = "Prior Distribution")
```

----

### Posterior Distribution for different Cauchy Scales

```{r fig.align='center', fig.width=10}
sigma_cauchy10 <- extract.samples(m8.2.cauchy.10, pars="sigma")
sigma_cauchy1 <- extract.samples(m8.2.cauchy.1, pars="sigma")
sigma_cauchy.point.1 <- extract.samples(m8.2.cauchy.point.1, pars="sigma")
dens(sigma_cauchy10[[1]], xlab="sigma", xlim=c(0.5,1.5), col="black")
dens(sigma_cauchy1[[1]], add=TRUE, col="blue")
dens(sigma_cauchy.point.1[[1]], add=TRUE, col="red")
legend(1.2, 6, legend=c("10", "1", "0.1"), col=c("black", "blue", "red"), lty=1:1, cex=0.8, title = "Posterior Distribution")
```

## A wild chain

This one will go nuts.

````
```{r}`r ''`
y <- c(-1,1)
m8.2 <- map2stan(
  alist(
    y ~ dnorm( mu , sigma ) ,
    mu <- alpha
    ) ,
  data=list(y=y) , start=list(alpha=0,sigma=1) ,
  chains=2 , iter=4000 , warmup=1000 )
```
````
----

### A wild chain's trace

- a problem with some models is that there are broad flat regions of the posterior density.
- often with flat priors
- can result in erratical samples

```{r fig.align='center', fig.height=3, fig.width=10}
y <- c(-1,1)
m8.2 <- readRDS("model/m8.2.rds")
plot(m8.2)
```

----

### Summary

```{r}
precis(m8.2)
```

Bringing in weak priors helps a lot.

```{r fig.height=4, fig.width=10}
m8.3 <- readRDS("model/m8.3.rds")
plot(m8.3)
```

----

### Weak Priors

Prior introduction

**alpha ~ dnorm( 1 , 10 ) ,**
**sigma ~ dcauchy( 0 , 1 )**

```{r}
precis(m8.3)
```

## Comparing different models Exercise 8H2

```{r}
data(WaffleDivorce)
d <- WaffleDivorce
d$MedianAgeMarriage_s <- (d$MedianAgeMarriage-mean(d$MedianAgeMarriage))/
  sd(d$MedianAgeMarriage)
d$Marriage_s <- (d$Marriage - mean(d$Marriage))/sd(d$Marriage)
df <- d[, c("Divorce", "MedianAgeMarriage_s", "Marriage_s")]
```

```{r}
m5.1_stan <- readRDS("model/m5.1_stan.rds")

m5.2_stan <- readRDS("model/m5.2_stan.rds")

m5.3_stan <- readRDS("model/m5.3_stan.rds")
```

```{r}
# compare the resulting models
compare(m5.1_stan,m5.2_stan,m5.3_stan)
```


## Thank you for your attention.















